---
title: "p8105_hw3_zl3544"
output: html_document
date: "2024-10-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lubridate)
library(ggplot2)
library(ggridges)
library(tidyverse)
library(p8105.datasets)
```

# Problem 1

## Exploration of the dataset
```{r}
# Display the structure of the dataset
str(ny_noaa)

# Get the number of rows and columns
nrow(ny_noaa)
ncol(ny_noaa)

# Display the dimensions of the dataset
dim(ny_noaa)

# Show the structure again for a detailed overview
str(ny_noaa)

# Provide a summary of the dataset
summary(ny_noaa)
```
### Description of the dataset

The dataset have 2595176 and 7 columns, which means that the ny_noaa dataset has 2,595,176 observations and 7 variables. The variables includes: id(Weather station identifier),date(Date of the observation), prcp(Precipitation (tenths of mm)), snow(Snowfall (mm)), snwd(Snow depth (mm)),tmax(Maximum temperature (tenths of degrees C)), tmin(Minimum temperature (tenths of degrees C)). In the summary, we can see that: prcp has a mean 29.82, the min is 0 and the max is 22860; snow has a mean 5, the min is -13 and the max is 10160; snwd has a mean 37.3, the min is 0 and the max is 9195. Also， there are missing value in some variables.

### Missing data

Caculate the percentage of missing data:
```{r}
ny_noaa %>%
  summarise(
    missing_prcp = mean(is.na(prcp)) * 100,
    missing_snow = mean(is.na(snow)) * 100,
    missing_snwd = mean(is.na(snwd)) * 100,
    missing_tmax = mean(is.na(tmax)) * 100,
    missing_tmin = mean(is.na(tmin)) * 100
  )

```
5.62% of prcp values are missing, 14.68% of snow values are missing,22.80% of snwd values are missing, 43.69% tmax are missing, 46.23% of tmin are missing.

## Data cleaning
```{r}
ny_noaa_cleaned= ny_noaa %>%
  janitor::clean_names() %>% 
  mutate(
    year = year(date),
    month = month(date),
    day = day(date),
    prcp = as.numeric(prcp)/10,
    tmax = as.numeric(tmax)/10,
    tmin = as.numeric(tmin)/10
  ) %>% 
  filter(tmax > tmin | is.na(tmax) | is.na(tmin)) %>% 
  filter(!(is.na(snow) & is.na(snwd) & is.na(tmax) & is.na(tmin) & is.na(prcp)))
head(ny_noaa_cleaned)
```
The data cleaning process: 

* Create separate variables for year, month, and day. 

* Ensure observations for temperature, precipitation, and snowfall are given in reasonable units.

* Ensures that only keep observations where the temperature data makes sense (i.e., maximum temperatures are logically higher than minimum temperatures)

* Filtered out rows in a dataset where all of the specified variables (snow, snwd, tmax, tmin, and prcp) are missing.

### Most common snowfall values:
```{r}
ny_noaa_cleaned %>% 
  count(snow) %>% 
  arrange(desc(n)) 
```
According to the table, 0 is the most common snowfall value.
